Fission product decay chain equilibria in \zirconia

### 2019-03-09 manuscript.tex
Update abstract

PCI failures in nuclear fuel claddings occur after power ramps, typically with a time delay of tens of minutes. The exact mechanism of failure is not known, however, it is widely believed that fission product SCC is involved in the process to some extent. The internal oxide layer of zirconium-based fuel claddings is protective, and so this layer must be breached before corrosive species can attack the metal underneath. Iodine is one of the main corrosive fission products produced in the fuel, though it is also produced in significant quantities by the decay of Te. Many Te fission product isotopes have half-lives under an hour and high fission yields, making them a large contributor to the iodine inventory during power ramps. Iodine nuclei are also unstable and may decay during the time delay into Xe, with the decay chain continuing to Cs over longer time spans. Fission predominantly occurs at the rim of the fuel pellet near the cladding, leading to fission product implantation within the oxide layer. It is important to determine the effect this can have on the oxide's protective capacity. We therefore examine the defect equilibria and report defect volumes of the fission products Te, I, Xe and Cs in the internal oxide layer of fuel cladding using a quantum mechanical modelling approach.

@@

With each nuclear decay comes a change in the chemical and therefore physical behaviour of the atom with its immediate environment. For example, an iodine dopant in \zirconia\ may decay into xenon, which may then have a significantly different thermodynamic equilibrium site preference from the one it inherited. Determining the behaviour of each of these elements in the oxide layer may provide information about the initiation of SCC in fuel cladding. We have therefore adopted a quantum-mechanical calculation approach to model the behaviour of the decay chain elements tellurium through caesium within tetragonal phase zirconia. \\

@@

Using these convergence criteria, non-defective supercells of tetragonal (3x3x2 unit cells) \zirconia\ were relaxed under constant pressure. The resulting structure was used as the starting point to which defects were introduced, and subsequently relaxed again, this time under constant volume conditions to simulate low defect concentrations \cite{Murphy2014, Bell2015}. The supercell was constructed such that its dimensions were as close to a cube as possible in order to minimise directional defect self-interaction across the periodic boundaries.

@@

\begin{figure}
    \centering
    \includegraphics[height=8cm]{img/te_decay_chain.png}
    \caption{Diagram of the Te$_{135}$ decay chain with decay modes and half-lives shown.}
    \label{fig:decaychain}
\end{figure}

@@

         \addplot[no marks, draw=red!60!yellow] table [x=pO2, y=Oi{-2},]{dat/cs_tet_10-5.dat}; \addlegendentry{\ch{O_{i}^{''}}};

@@

\begin{table}
\caption{Independent fission product yields and half-lives for the major iodine isotopes and precursors in a thermal neutron reactor. Yields from the Joint Evaluated Fission and Fusion File (JEFF 3.3). All isotopes undergo single $\beta$- decay. Metastable states are included.}  \label{table:decaydata}
\begin{center}
\begin{tabular}{c c c}
\hline
Isotope & Independent Yield (\%) & Half-life \\
\hline
\texorpdfstring{I\textsubscript{131}}{I131} & 0.0029604
 & 8.023 d \cite{I131halflife} \\
\texorpdfstring{Te\textsubscript{132}}{Te132} & 1.5042 & 3.180 d \cite{Te132} \\
\texorpdfstring{I\textsubscript{132}}{I132} & 0.028399  & 2.295 h \cite{Te132} \\
\texorpdfstring{Te\textsubscript{133}}{Te133} & 3.9386 & 12.5 m \cite{khazov2011nuclear} \\
\texorpdfstring{I\textsubscript{133}}{I133} & 0.198194  & 20.87 h \cite{I133} \\
\texorpdfstring{Te\textsubscript{134}}{Te134} & 6.3001 & 41.8 m \cite{Sonzogni2004} \\
\texorpdfstring{I\textsubscript{134}}{I134} & 0.7673 & 52.5 m \cite{Sonzogni2004} \\
\texorpdfstring{Te\textsubscript{135}}{Te135} & 3.4777 & 19.0 s \cite{tellurium135halflife} \\
\texorpdfstring{I\textsubscript{135}}{I135} & 2.4635 & 6.58 h \cite{Singh2008} \\ 
\texorpdfstring{Te\textsubscript{136}}{Te136} & 1.6652 & 17.63  s \cite{Mccutchan2018NuclearA=136} \\
\texorpdfstring{I\textsubscript{136}}{I136} & 3.03823 & 83.4 s \cite{Mccutchan2018NuclearA=136} \\
\texorpdfstring{Te\textsubscript{137}}{Te137} & 0.48440 & 2.49 s \cite{browne2007nuclear} \\
\texorpdfstring{I\textsubscript{137}}{I137} & 2.7026 & 24.5 s \cite{browne2007nuclear} \\ 
\texorpdfstring{Te\textsubscript{138}}{Te138} & 0.11280 & 1.4 s \cite{chen2017nuclear} \\ 
\texorpdfstring{I\textsubscript{138}}{I138} & 1.2419 & 6.26 s \cite{chen2017nuclear} \\  \hline
\end{tabular}
\end{center}
\end{table}

>>>>

The defect equilibria of fission products Te, I, Xe and Cs and their defect volumes are predicted in tetragonal \zirconia\ using density functional theory calculations. Dopant fission product concentrations of 10$^{-5}$ to 10$^{-3}$ atoms per formula unit of \zirconia\ are examined. These data are presented in Brouwer diagrams to show the calculated defect concentrations between oxygen partial pressures of 10$^{-35}$ and 10$^{0}$ atm. The dominant Te defects in \zirconia\ are \ch{Te^{**}_{O}} and \ch{Te^{'''}_{Zr}}, with defect volumes relative to the non-defective crystal of 2.74 \r{A}$^{3}$ and 59.93 \r{A}$^{3}$, respectively. Upon $\beta$- decay, these defects become \ch{I^{*}_{O}} and \ch{I^{'''}_{Zr}}, with defect volumes of 29.90 \r{A}$^{3}$ and 87.07 \r{A}$^{3}$, respectively. Upon a second $\beta$- decay, \ch{Xe^{**}_{O}}, \ch{Xe^{'''}_{Zr}} and \ch{Xe^{''''}_{Zr}} are produced, with defect volumes of 9.02 \r{A}$^{3}$, 91.65 \r{A}$^{3}$ and 115.50 \r{A}$^{3}$ respectively. The large positive volumes of these defects generate stresses in the \zirconia\ lattice, and if produced in significant quantities (such as during a power ramp), could promote crack formation. PCI failures, as reported in the literature, exhibit time delays between reactor power ramping and failure which are in line with the decay rate of Te and I isotopes. This has implications for fuel and cladding designs which are resistant to SCC and PCI.

%PCI failures in nuclear fuel claddings occur after power ramps, typically with a time delay of tens of minutes. The exact mechanism of failure is not known, however, it is widely believed that fission product SCC is involved in the process to some extent. The internal oxide layer of zirconium-based fuel claddings is protective, and so this layer must be breached before corrosive species can attack the metal underneath. Iodine is one of the main corrosive fission products produced in the fuel, though it is also produced in significant quantities by the decay of Te. Many Te fission product isotopes have half-lives under an hour and high fission yields, making them a large contributor to the iodine inventory during power ramps. Iodine nuclei are also unstable and may decay during the time delay into Xe, with the decay chain continuing to Cs over longer time spans. Fission predominantly occurs at the rim of the fuel pellet near the cladding, leading to fission product implantation within the oxide layer. It is important to determine the effect this can have on the oxide's protective capacity. We therefore examine the defect equilibria and report defect volumes of the fission products Te, I, Xe and Cs in the internal oxide layer of fuel cladding using a quantum mechanical modelling approach.

@@

With each nuclear decay comes a change in the chemical and therefore physical behaviour of the atom with its immediate environment. For example, an iodine dopant in \zirconia\ may decay into xenon, which may then have a significantly different thermodynamic equilibrium site preference from the one it inherited. This phenomenon is known as "radioparagenesis", and was first considered in the context of nuclear waste disposal \cite{Jiang2009PredictingBaCl, Jiang2010UsingForms}. Determining the behaviour of each of these elements in the oxide layer may provide information about the initiation of SCC in fuel cladding. We have therefore adopted a quantum-mechanical calculation approach to model the behaviour of the decay chain elements tellurium through caesium within tetragonal phase zirconia. \\

@@

Using these convergence criteria, non-defective supercells of tetragonal (3x3x2 unit cells) \zirconia\ were relaxed under constant pressure. The resulting structure was used as the starting point to which defects were introduced, and subsequently relaxed again, this time under constant volume conditions to simulate low defect concentrations \cite{Murphy2014, Bell2015}. The supercell was constructed such that its dimensions were as equal as possible in all directions in order to minimise directional defect self-interaction across the periodic boundaries.

@@

\begin{figure}
    \centering
    \includegraphics[height=8cm]{img/te_decay_chain2.png}
    \caption{Diagram of the Te$_{135}$ decay chain with decay modes and half-lives shown.}
    \label{fig:decaychain}
\end{figure}

@@

         %\addplot[no marks, draw=red!60!yellow] table [x=pO2, y=Oi{-2},]{dat/cs_tet_10-5.dat}; \addlegendentry{\ch{O_{i}^{''}}};

@@

\begin{table}
\caption{Independent fission product yields and half-lives for the major iodine isotopes and precursors in a thermal neutron reactor. Yields from the Joint Evaluated Fission and Fusion File (JEFF 3.3). All isotopes undergo single $\beta$- decay. Metastable states are included.}  \label{table:decaydata}
\begin{center}
\begin{tabular}{c c c}
\hline
Isotope & Independent Yield (\%) & Half-life \\
\hline
\texorpdfstring{I\textsubscript{131}}{I131} & 0.0029604
 & 8.023 d \cite{I131halflife} \\
\texorpdfstring{Te\textsubscript{132}}{Te132} & 1.5042 & 3.180 d \cite{Te132} \\
\texorpdfstring{I\textsubscript{132}}{I132} & 0.028399  & 2.295 h \cite{Te132} \\
\texorpdfstring{Te\textsubscript{133}}{Te133} & 3.9386 & 12.5 m \cite{khazov2011nuclear} \\
\texorpdfstring{I\textsubscript{133}}{I133} & 0.198194  & 20.87 h \cite{I133} \\
\texorpdfstring{Te\textsubscript{134}}{Te134} & 6.3001 & 41.8 m \cite{Sonzogni2004} \\
\texorpdfstring{I\textsubscript{134}}{I134} & 0.7673 & 52.5 m \cite{Sonzogni2004} \\
\texorpdfstring{Te\textsubscript{135}}{Te135} & 3.4777 & 19.0 s \cite{Singh2008} \\
\texorpdfstring{I\textsubscript{135}}{I135} & 2.4635 & 6.58 h \cite{Singh2008} \\ 
\texorpdfstring{Te\textsubscript{136}}{Te136} & 1.6652 & 17.63  s \cite{Mccutchan2018NuclearA=136} \\
\texorpdfstring{I\textsubscript{136}}{I136} & 3.03823 & 83.4 s \cite{Mccutchan2018NuclearA=136} \\
\texorpdfstring{Te\textsubscript{137}}{Te137} & 0.48440 & 2.49 s \cite{browne2007nuclear} \\
\texorpdfstring{I\textsubscript{137}}{I137} & 2.7026 & 24.5 s \cite{browne2007nuclear} \\ 
\texorpdfstring{Te\textsubscript{138}}{Te138} & 0.11280 & 1.4 s \cite{chen2017nuclear} \\ 
\texorpdfstring{I\textsubscript{138}}{I138} & 1.2419 & 6.26 s \cite{chen2017nuclear} \\  \hline
\end{tabular}
\end{center}
\end{table}
